# Basics {#basics}

```{r child="work-in-progress.qmd"}
```

```{r components, include = FALSE}
eunit1 <- function() '<i class="fas fa-seedling"></i>'
eunit2 <- function() '<i class="fas fa-wine-bottle"></i>'
eunit3 <- function() '<i class="fas fa-laptop"></i>'
```

::: {.blockquote}
If the design of an experiment is faulty, any method of interpretation which makes it out to be decisive must be faulty too.

--Fisher, 1935, Design of Experiments
:::

There are many articles and books dedicated to explaining the concepts of experimental designs. This includes the seminal book by @Fisher1935-qc, @Bailey2008-gw, etc. This chapter presents an overview of the statistical concepts that are most pertinent in designing an experiment. 

Most experiments are comparative in nature, or more specifically, most experiments involve study of two or more experimental conditions in which the primary interest is to compare the outcome under the different conditions. You'll therefore find that most experiments in this book are **comparative experiments**. 

At the heart of each experiment, we are ultimately seeking confidence in any conclusion we are making from the analysis of the experimental data. What is perhaps not emphasised enough is that the **_experiments are human endeavours_**. Often there are multiple people involved in running experiments and a key challenge is to ensure each individual has sufficient understanding to play their role well.  We touch more on this in Chapter -@management. 

All experiments have a cost whether that be financial, resources, time or other. We can consider the **experimental cost** as a function of the ability to redo the experiment again. For example, if your experimental resources are based on examination of fossils and the fossils are destroyed in the experimental process, then the cost of the experiment is infinite -- you only have one chance to do the experiment -- so it's absolutely essential that you plan, design and execute the experiment well.

Doing an experimental well requires a good understanding of the subject matter so that potential sources of variation can be controlled or accounted for in the experimental design. Prior to the collection of data, the statistical component of an experiment tends to be focussed on the design of the experiment, i.e. how the treatments are assigned to experimental units under practical constrains to maximise the statistical information of interest. Under this consideration, the statistical problem for experimental design is often reduced to either a randomisation or an optimisation problem and the experimental context may be stripped away in the generation of the design using computer software. We discuss this in Chapter @context, describe a system that  encourages higher order thinking of the experimental design, termed the grammar of experimetnal designs, in Chapter @grammar, implemented as the `edibble` R-package and present how to get started with the `edibble` system to construct experimental design in Chapter @start.

Before we get into the crux of the basics of the experimental design, consider the three scenarios below. Each scenario describes an experiment where technical details have been reduced so it doesn't serve as a distraction for now. For each scenario, try to see if you can identify what are the basic components to build the design of the experiment.
 
::: {.callout-note}

## Scenario 1: plant growth  `r eunit1()` 

Microbes have a potential for promoting plant growth by increasing abiotic stress tolerance of the plant. An experiment is conducted to study four bacterial strains known for promoting plant growth under osmotic stress.

:::

::: {.callout-note}

## Scenario 2: wine tasting `r eunit2()`

A large-scale experiment was conducted to discern whether expensive wines are good value for money. The experiment involved 450 individuals in a public tasting of expensive or inexpensive wine. The taster did not know whether they were tasting the expensive or inexpensive wine. 

:::

::: {.callout-note}

## Scenario 3: new classification method `r eunit3()`

A new statistical method is proposed as a better alternative for classification of cellular type from single cell transcriptomics data. A number of simulation, based on a number of public benchmark  datasets, was carried out to compare the method under different metrics. 

:::
 

## Basic terminology {#terminology}

The field of experimental design is large and is applied to many domains. While some specialised terminology exist, the following terminology are basic terms that are commonly understood across domains:

* **treatments**: the whole set of experimental conditions in which the units are subjected to,
* **experimental unit**: the smallest unit in which the treatment can independently be applied,
* **observational unit**: the smallest unit in which the response is measured on,
* **block** or **cluster**: the grouping of another unit such that the units within the same block or cluster is more alike in some aspect than units in different blocks or clusters,
* **design**: the allocation of treatment to units, 
* **plan** or **layout**: the actual assignment of treatment to units, and
* **experimental structure**: the arrangements or constrains of experimental factors -- the structure can be broadly categorised as: 
  * **unit structure**: the relation between the units, e.g. pots nested in particular glasshouses, subjects segregated by sex, and so on; and
  * **treatment structure**: the division of treatment levels to meaningful groups, e.g. factorial structure means all combinations across the different treatment factors.
  
The above terminology describe the experimental factors and its associations. More specifically, **experimental factors** are any variables that pertain to the experiment (physical or otherwise) and includes the treatments, experimental units, observational units and blocks/clusters. The associations of experimental factors include its nesting structure or say the assignment of treatment to units. 

In addition to above, there are commonly understood terminology that refer to the properties of experimental designs. These include: 

* **balanced experiment**: refers to when the replication of the treatments are equal,
* **orthogonal design**
* **complete design**: every treatment appears exactly once in each block or cluster.

::: {.callout-note}

## plant growth

* The experimental aim is to study the effects of plant growth of four particular bacterial strains. 
* The experimental units are ... 
:::


### Specialised terminology

Experimental design is applicable across numerous domains, so perhaps it is not surprising that there are domain specific or specialised terminology that is used to describe the whole experimental designs, e.g. Latin Square Design, Step-Wedge Trial, Beehive Design and so on. Some recipe of these named experimental design are presented and discussed in Chapter @cookbook.


## Principles

All experiments have a potential to go wrong -- plans may not be followed to the letter; you can't distinguish which factor had the effect; or unknown factor is affecting the response. The degree of how wrong it went can vary but in some cases you wouldn't even know it went wrong!

In designing an experiment, there are some safety measures that you can put in place to minimise potential issues. The basic statistical principles of experimental design outlined below form some of these safety measures.  

### Randomisation and optimisation

**Randomisation** in the context of experimental design means that the treatment allocation to units were probabilistically determined by a mechanism where the outcome is not (consciously or unconsciously) influenced by those involved in the experiment. The opposite spectrum of randomisation is **systematic** allocations of treatments to units -- this means that the allocations are not probabilistic and if you know the systematic order, then you can precisely infer the allocation.  

Randomisation is a core principle of experimental design. When you conduct an experiment, you are investigating factors to understand some process or phenomenon, and there will be factors which affect the response that you don't know about. 

Not all experiments can allocate treatments to units randomly, e.g. you can't get your participants to smoke to study the effect of smoking, nor can you get a random agricultural land to have a drought to study how well a crop grows. These types of experiments are referred to as *quasi-experiment* or *natural experiment*. 




### Replications, repetition and duplication


### Pseudo-replications

### Confounding

## Summary

The major concepts this chapter covered are:

* basic statistical terminology for experimental design 
* importance of replication and randomisation in experiments 
* pitfalls of pseudo-replication 

## Exercises

1. A farmer would like to know which diet is most effective in increasing the weight of the pigs. There are three types of diet that he wishes to test. For this scenario, describe the experimental aim, treatments, experimental units, observational units and the experimental design.
