# Terminology {#sec-terminology}

```{r}
#| child: "status.qmd"
```
```{r}
#| include: false
source("setup.R")
```

The field of experimental design is large and is applied to many domains, e.g. biology, psychology, industry, finance, marketing and so on. While some specialised terminology exist, the following terminology are basic terms that are commonly understood across domains:



## Unit related 

* **experimental unit**: the smallest unit in which the treatment can _independently_ be applied,
* **observational unit**: the smallest unit in which the response is measured on,
* **block** or **cluster**: the grouping of another unit such that the units within the same block or cluster is more alike in some aspect than units in different blocks or clusters,
* **unit structure**: the relation between the units, e.g. pots nested in particular glasshouses, subjects segregated by sex, and so on


## Treatment related 

* **treatments**: the whole set of experimental conditions in which the units are subjected to,
* **treatment structure**: the division of treatment levels to meaningful groups, e.g. factorial structure means all combinations across the different treatment factors,
* **control** in treatment refers to a standard approach used, e.g. chemotherapy is a standard way to treat patients with particular cancers, and
* **placebo** means a treatment designed to 



## Experiment related 

* **design**: the allocation of treatment to units, 
* **plan** or **layout**: the actual assignment of treatment to units, and
* **experimental structure**: the arrangements or constrains of experimental factors -- the structure can be broadly categorised as unit and treatment structure. 
  

  
The above terminology describe the experimental factors and its associations. More specifically, **experimental factors** are any variables that pertain to the experiment (physical or otherwise) and includes the treatments, experimental units, observational units and blocks/clusters. The associations of experimental factors include its nesting structure or say the assignment of treatment to units. 

In addition to above, there are commonly understood terminology that refer to the properties of experimental designs. These include: 

* **balanced experiment**: refers to when the replication of the treatments are equal,
* **orthogonal design**
* **complete block design**: every treatment appears exactly once in each block or cluster, otherwise it is referred to as **incomplete block design**,
* **blinded experiments**, relevant only to studies involving human subjects, refers to experiments where subjects are not informed about which treatment they received. **Double-blinded** studies means that both those administrating and receipient of the treatment are not informed about the actual treatment. 

::: {.callout-note icon=false}

## `r eunit1()`  plant growth

* The experimental aim is to study the effects of plant growth of four particular bacterial strains. 
* The experimental units are ... 
:::


## Specialised terminology

Experimental design is applicable across numerous domains, so perhaps it is not surprising that there are domain specific or specialised terminology that is used to describe the whole experimental designs, e.g. Latin Square Design, Step-Wedge Trial, Beehive Design and so on. Some recipe of these named experimental design are presented and discussed in [Chapter -@sec-cookbook].