# Completely randomised design 

```{r}
#| child: "status.qmd"
```
```{r setup, include = FALSE}
source("setup.R")
set.seed(1)
```

A completely randomised design, commonly abbreviated as CRD, contain two factors: experimental units and treatments with a completely unstructured experiment (i.e. no grouping for the factors). CRD is sufficiently characterised by the number of units, $n$, and the number of treatment, $t$. Instead of $n$, you ccan parameterise the design with the number of replications, $r$; in which case, you can derive $n = rt$. The later parameterisation means that the design will be balanced while the former does not guarantee that the design is balanced. 

You can call on `menu_crd()` to see the code in the terms of the fundamental system. If you omit the argument values, then a random parameterisation will be selected for you. You can find more details about the design in the documentation (`?menu_crd`).

```{r, comment = ""}
menu_crd()
```

If you want to generate the design table, then you can use the `takeout()` function parsing the named design object created by the set of `menu_` functions. If you don't select any menu then the `takeout()` function will automatically select a random menu for you. You can find the list of available named experimental designs by calling on `scan_menu()`. For now, let's "takeout" the completely randomised design with 5 treatments and 30 experimental units.


```{r}
crd <- takeout(menu_crd(t = 5, n = 30, seed = 1), )
crd
```

A design table generated by `takeout()` is an object with a special class of `edbl_table`. When you print a `takeout` object, a table will be displayed like a typical `edbl_table` object but the recipe code will also be displayed prior to the table. 


You can quickly plot the design using the `autoplot()` in the `deggust` package. 

```{r autoplot-crd, fig.height = 4, fig.width = 5}
autoplot(crd)
```
