# Grammar {#sec-grammar .unnumbered}

```{r}
#| child: "status.qmd"
```

```{r setup, include = FALSE}
source("setup.R")
```

::: blockquote

Grammar  makes  language  expressive.  A  language  consisting  of  words and no grammar (statement = word) expresses only as many ideas as there are words. By specifying how words are combined in statements, a grammar expands a language’s scope.

-- Wilkinson, 2005, The Grammar of Graphics, 2nd edition

:::




In this chapter, I describe an object-oriented programming system called "the grammar of experimental designs" that is designed to capture elements of the origin of an experimental data encapsulated in a cognitive framework. The grammar of experimental designs is a term that is heavily influenced by @Wilkinson2005-oz, who use the term "grammar of graphics" to describe an object-oriented system to create unlimited number of plots from following a relatively small number of rules. The grammar of graphics is adopted across many systems (Julia, Python, R, Tableau and so on) with its most popular interpretation [@Wickham2010-kt] as the `ggplot2` R-package [@ggplot2]. An overview of `ggplot2` is given in @sec-ggplot. The grammar of experimental designs leverage aspects of the grammar of graphics for the visualisation of the experimental designs as discussed more extensively in @sec-deggust. 

Henceforth, "grammar" refers to the grammar of experimental designs unless explicitly stated otherwise.





