# Visualise {#sec-deggust}

```{r}
#| child: "status.qmd"
```

```{r setup, include = FALSE}
source("setup.R")
```

Visualisation of the experimental design is helpful for diagnostics (as outlined in @sec-dignosis), however, it serves well for communication purposes. There are broadly four types of purposes in visualising experimental design as outlined in @tbl-purpose. 


\# | Purpose | Primary audience | Output requirement
-: | -------- | --- | -------
1 | Checking the experimental structure | Domain experts | Quick and easy 
2 | Diagnosis of experimental design | Statisticians | Informative
3 | Implementing the experimental design | Technicians | Detailed diagram
4 | Publications | Public | Polished 

: Purposes of visualising experimental designs {#tbl-purpose}



## Rectangular display 

```{r dummy-crd, echo = FALSE}
crd_eg <- takeout(menu_crd(t = 4, r = 6))
knitr::kable(crd_eg)
```

```{r dummy-crd-plot, echo = FALSE}
ggplot(crd_eg, aes(unit, "1", color = trt)) +
  geom_point(size = 7) +
  theme_void()
```


```{r dummy-crd-autoplot, echo = FALSE}
autoplot(crd_eg) +
  theme(plot.title = element_blank(),
        plot.subtitle = element_blank())
```

