# Practice {#sec-practice .unnumbered}


```{r}
#| child: "status.qmd"
```

```{r setup, include = FALSE}
source("setup.R")
```

::: blockquote

... communication is complex, fraught with tensions, misunderstandings, and problems — rather than a simple process of creating shared meaning.

-- Littlejohn et al., 2017, Theories of Human Communication

:::

Designing an experiment _in practice_ isn't just about applying statistical principles as covered in @sec-principles. In fact, a large component of experimental design in practice is being able to _elicit information from the experts and translating it to the statistical structure needed for the experimental design_. Therefore, any best practice for experimental design is incomplete without discussing the practical factors that comes into play in designing an experiment.  

This part is organised as follows:

* @sec-consultation focuses on consulting aspects from the perspective of the statistician,
* @sec-management discusses the project management across all parties involved, 
* @sec-context considers the domain knowledge aspects and some context specific examples, and
* @sec-guides present guides for general consideration of experimental design.

