# Preface {-}

```{r setup, include = FALSE}
source("setup.R")
knitr::read_chunk("setup.R")
```

```{r child="work-in-progress.qmd"}
```


This book is intended for anyone who is interested in planning, designing and simulating experiments with a programming language. The programming is illustrated with the R language, however, the concept may be generalised to other programming languages. 

The book can also serve as a reference to both undergraduate and graduate courses. Each section has problems that can be used as exercise for students. The book can also be used by practitioners that are involved with scientific or industrial experiments. The book will include plenty of examples spanning a diverse range of disciplines.

It is beneficial to have at least some knowledge of R. Readers may like to read XXX for basic knowledge in R and "R for Data Science" for `tidyverse` approaches to data analysis in R.  

The book primarily makes use of `ggplot2` for drawing static plots. The default color scheme and theme are modified as below and are applied for all plots produced by `ggplot2` henceforth.

```{r ggplot2}
```
